Flat histogram method comparison on 2D Ising Model

### 2019-07-03 papers/histogram/ising.tex
working on ising paper currently on sad section.

Liang independently considered whether WL could be treated as a special case of
Stochastic Approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form of
Stochastic Approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte-Carlo
methods across arbitrary systems.  The SAD (Stochastic Approximation with
Dynamic $\gamma$) method~\jpsays{outlined by Pommerenck et.al include David and
my paper on SAD square well in citations!} is a special version of the SAMC
algorithm that dynamically chooses the modification factor rather than relying
on system dependent parameters. SAD shares the same convergence properties with
SAMC while replacing un-physical user-defined parameters with the algorithms
dynamic choice.  In this work, we apply the family of weight-based flat
histogram Monte Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

@@

The $1/t$-WL method is a refinement of the original implementation that
addresses the error saturation of WL. Belardinelli and Pereyra
implemented a schedule~\cite{belardinelli2007fast} that enforced that
if $\gamma < \nicefrac{1}{t}$ then the update factor is set to
$\nicefrac{1}{t}$ and the histogram is no longer tracked. Using this
metric, the error saturation is avoided since the correct density of
states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}.

\jpsays{I am here!}

Schneider $\text{et. al}$ outline minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL to
$1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. As per the original
$1/t$ implementation~\cite{belardinelli2007fast}, the update factor is
decreased once all $N_S$ energy states are visited at least once (i.e.
$H(E) \neq 0$) effectively avoiding the concept of `flatness'.  In this
work, we refer to this refined algorithm as $1/t$-WL, with the update
factor asymptoting to $\nicefrac{N_S}{t}$.

\subsection{SAMC}
The Stochastic Approximation Monte Carlo (SAMC) algorithm has a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.
SAMC offers extreme simplicity and proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  Unlike WL methods, the energy range need not be
known \emph{a priori.}  The convergenc time depends only on the choice of
parameter $t_0$.  Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}.

\subsection{SAD Algorithm}\label{sec:sad}
In our previous work we discuss this in detail so here we only highlight...
Stochastic Approximation w/ Dynamic $\gamma$ here.

@@

\section{Acknowledgements}

>>>>

Liang independently considered whether WL could be treated as a special case of
Stochastic Approximation whose convergence could be mathematically
proven~\cite{liang2006theory, liang2007stochastic}. In 2007, Liang et
al.~\cite{liang2007stochastic} argued that WL can be considered a form of
Stochastic Approximation Monte Carlo (SAMC). Unlike WL, SAMC can guarantee
convergence (if certain conditions are met). Despite the added benefit of
guaranteed convergence, the method still has a system specific user-defined
variable. Such variables often create difficulty when applying Monte-Carlo
methods across arbitrary systems.  The SAD (Stochastic Approximation with
Dynamic $\gamma$) method outlined by Pommerenck
et.al~\cite{pommerenck2019stochastic} is a special version of the SAMC algorithm
that dynamically chooses the modification factor rather than relying on system
dependent parameters. SAD shares the same convergence properties with SAMC while
replacing un-physical user-defined parameters with the algorithms dynamic
choice.  In this work, we apply the family of weight-based flat histogram Monte
Carlo methods (WL, 1/t-WL, SAMC, SAD) to the 2D Ising model.

@@

The $1/t$-WL method is a refinement of the original implementation that
addresses the error saturation of WL. Belardinelli and Pereyra
implemented a schedule~\cite{belardinelli2007fast} that enforced that
if $\gamma < \nicefrac{1}{t}$ then the update factor is set to
$\nicefrac{1}{t}$ and the histogram is no longer tracked. Using this
metric, the error saturation is avoided since the correct density of
states is approached asymptotically as
$t^{-\frac12}$~\cite{belardinelli2008analysis}.
Schneider $\text{et. al}$ outline minor refinements to the $1/t$-WL algorithm
including $\nicefrac{N_S}{t}$ scaling and switching from standard WL to
$1/t$-WL when the update factor $\gamma <
\nicefrac{N_S}{t}$~\cite{schneider2017convergence}. As per the original
$1/t$ implementation~\cite{belardinelli2007fast}, the update factor is
decreased once all $N_S$ energy states are visited at least once (i.e.
$H(E) \neq 0$) effectively avoiding the concept of `flatness'.

Another weight-based flat histogram method is
the Stochastic Approximation Monte Carlo (SAMC) algorithm. SAMC has a simple
schedule by which the update factor $\gamma$ is continuously
decreased~\cite{liang2007stochastic, werlich2015stochastic,
schneider2017convergence}.  The update factor is defined in the
original implementation~\cite{liang2007stochastic} in terms of an
arbitrary tunable parameter $t_0$,
\begin{align}
\gamma_{t}^{\text{SA}} =\frac{t_0}{\max(t_0,t)}
\end{align}
where as above $t$ is the number of moves that have been attempted.

SAMC offers extreme simplicity and proven convergence.
Provided the update factor satisfies
\begin{align}
\sum_{t=1}^\infty \gamma_{t} = \infty \quad\textrm{and}\quad
\sum_{t=1}^\infty \gamma_{t}^\zeta < \infty
\end{align}
where $\zeta \in \{1,2\}$, Liang has shown that the weights are proven
to converge to the true density of states~\cite{liang2006theory,
liang2007stochastic}.  Unlike WL methods, the energy range need not be
known \emph{a priori.}  The convergenc time depends only on the choice of
parameter $t_0$.
The parameter $t_0$ can unfortunately be difficult to chose in advance
for arbitrary systems.
Liang \emph{et al.} give a rule of thumb in
which $t_0$ is chosen in the range from $2N_S$ to $100N_S$ where $N_S$
is the number of energy bins~\cite{liang2007stochastic}.  Schneider
\emph{et al.} found that for the Ising model this heuristic is helpful
for small spin systems, but that larger systems require an even higher
$t_0$ value~\cite{schneider2017convergence}. Our work also confirms this.

Pommerenck et al. propose a refinement~\cite{pommerenck2019stochastic} to SAMC where
the update factor is determined dynamically rather than by the user.
Stochastic Approximation with a dynamic $\gamma$ (SAD) requires the user to provide the
lowest temperature of interest $T_{\min}$. This is analogous to WL requiring a priori
an energy range of interest; however, this may easier to identify and is more physical
than the SAMC parameter $t_0$.
The update factor can be written in terms of the current estimates for the highest $E_H$ and
lowest $E_L$ energies of interest.
\begin{align}
  \gamma_{t}^{\text{SAD}} =
     \frac{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{t_L}
     }{
       \frac{E_{H}-E_{L}}{T_{\text{min}}} + \frac{t}{N_S}\frac{t}{t_L}
     }
\end{align}

@@
